---
title: "A quick introduction to crossr"
author: "Otho Mantegazza"
date: "`r Sys.Date()`"
output:
  knitr:::html_vignette:
    toc: yes
    number_sections: true
    theme: flatly
    highlight: default
vignette: >
  %\VignetteIndexEntry{overview of crossr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo=FALSE, message=FALSE}
library(crossr)
```


# Introduction

Crossr contains a set of functions and a workflow that I use to compare gene expression levels between species. 

To achieve this, crossr relies extensively on exploratory data analysis methods and not on strict statistical tests. Designing a statistical test for differential expression among species is very difficult because of the complexity of the underlying data and because of the low replicate number that we generally use in transcriptomic. 

Providing a full statistical test for differential expression among species is beyond the scope of this package and beyond my capability. What crossr provides instead is a simple explorative workflow that will allow you to make useful observation of how groups of homologous genes behave in different species.

Crossr builds on the output of :

- [Salmon](https://combine-lab.github.io/salmon/) for transcript quantification,
- [Orthofinder](https://github.com/davidemms/OrthoFinder) to determine homology groups between the species.

Here I will illustrate the workflow of crossr on the [GSE69077](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE69077) dataset,  which was published by [Rawat et al. in 2015](https://www.ncbi.nlm.nih.gov/pubmed/26382944). This dataset contains matched RNA-seq samples of heat-shocked and recovered *Arabidopsis thaliana* and *Arabidopsis lyrata* plants. This experimental setup perfectly matches what crossr is designed for.

In order to make this vignette lighter, I will use only a random subset of 500 orthogroups from original dataset. The sample dataset is available in the `exdata` folder.



# Read Mapping

Mapping of RNA-seq reads is not directly a part of the crossr workflow. Crossr requires only two normalized gene expression matrices (one for each species taken into consideration) as starting point.

For mapping of RNA-seq read we suggest **[Salmon](https://combine-lab.github.io/salmon/)** because it is fast, precise and easy to use. We also provide the `make_TPM_df()` function that parses and loads the Salmon output directly into R. 

These are the **bash** commands that we use to quantify  RNA-seq reads with Salmon (at this moment we have tested the 0.8.2 version):

```{bash, eval = FALSE}
## make index
./Salmon-0.8.2_linux_x86_64/bin/salmon\
  index -t path/to/genome.fasta\
  -i th_transcripts_index\
  --type quasi -k 31


## Quantify and save mapping to SAM files for diagnostics
for f in reads/*.fastq; do 

base=${f%.fastq}
echo $base
base=${base#reads/}
echo $base
outsam=path/to/sam/folder/$base.sam
echo $outsam

./Salmon-0.8.2_linux_x86_64/bin/salmon quant \
  -i path/to/index -l U \
  -r $f -o $base\
  -p 8\
 --writeMappings=$outsam
done
```


# Load mappping into R

You will have to load into R: transcript quantifications, and sample metadata.

## Load transcript quantification from Salmon
 
You can load the output of Salmon into R with the function `make_TPM_df()`. This function takes as input the **path** to the folder that the quantification output from Salmon.

In detail, the function `make_TPM_df()`:

1. extracts the TPM values from all the `quant.sf` files in every subfolder of the input path,
2. organizes those values in a `data.frame` using the **subfolder names** as column ids and the feature names as row ids. This `data.frame` is the **expression matrix**.

It does not matter it the subfolders contain additional files because `make_TPM_df()` will ignore them.

In our example, the folder that contains the quantification for *A. thaliana* has this structure:

```
bash> tree -A inst/exdata/thaliana

thaliana
├── SRR2033948
│   └── quant.sf
├── SRR2033949
│   └── quant.sf
├── SRR2033950
│   └── quant.sf
├── SRR2033951
│   └── quant.sf
├── SRR2033952
│   └── quant.sf
└── SRR2033953
    └── quant.sf

```

With `make_TPM_df` we can extract the TPM values from all the `quant.sf` files and build the expression matrix.

```{r}
##  get the path to the thaliana folder
thaliana_path <- system.file("exdata/thaliana", package = "crossr") 

## extract TPMs and build dataframe
thaliana <- make_TPM_df(thaliana_path)

## this is how the top rows of the dataframe should look like
head(thaliana)
```

And the same for ***A. lyrata***

```{r}
lyrata_path <- system.file("exdata/lyrata", package = "crossr")
lyrata <- make_TPM_df(lyrata_path)
str(lyrata)
```

While we suggest Salmon, you can use crossr downstream of **any mapping algorithm** but you will have to load the expression matrix in R by yourself. The package [tximport](http://bioconductor.org/packages/release/bioc/html/tximport.html) can help.

We anyway suggest to use expression values that:

- are normalized for library size,
- do not depend on gene length,

for example transcript per million (TPM) is a good option.

## Load sample metadata

In the expression matrix every column represents an RNAseq sample.

Usually **sample metadata** are available they might be used to change the `colnames` of the expression sets from the sample ID to names that provide more immediate information. In this example we can use this [sample sheet](https://www.ncbi.nlm.nih.gov/Traces/study/?acc=SRP058527&go=go) to do so.

This code:

- loads the feature table into R,

```{r}
sample_info_path <- system.file("exdata",
                                "SRP058527_metadata.txt",
                                package = "crossr")

sample_info <-  read.table(sample_info_path,
                           header = TRUE,
                           sep = "\t")  

```

- simplify the sample info so that they can be used as column name,

```{r}
reduce_info <- function(info) {
    if(info == "heat stress 0 h") "0h"
    else if(info == "heat stress 6 h 37\302\260C") "heat_6h"
    else "recovery"
}
sample_info$etime <- vapply(sample_info$exposure_time_s,
                            reduce_info,
                            character(1))
sample_info$etime <- paste(sample_info$etime,
                           c(1, 2),
                           sep = "_")
```

- and set the sample info as column name of both datasets instead of the SRR id.

```{r}
switch_name <- function(name) {
    if(name %in% sample_info$Run_s) {
    sample_info[sample_info$Run_s == name, "etime", drop = TRUE]
    } else name
    }
colnames(thaliana) <- vapply(colnames(thaliana), switch_name, character(1))
colnames(thaliana) <- paste("thal",
                            colnames(thaliana),
                            sep = "_")
str(thaliana)

colnames(lyrata) <- vapply(colnames(lyrata), switch_name, character(1))
colnames(lyrata) <- paste("lyr",
                          colnames(lyrata),
                          sep = "_")
str(lyrata)
```


# Homology groups

The homology groups are called orthogroups in [Orthofinder](https://github.com/davidemms/OrthoFinder) and are defined as:

> The set of genes that are descended from a single gene in the last common ancestor of all the species being considered.

## Ortholog inference

We use **Orthofinder** for ortholog inference. Discussing Orthofinder is beyond the scope of this document; please, refer to its [documentation](https://github.com/davidemms/OrthoFinder/blob/master/OrthoFinder-manual.pdf) for details.

Orthofinder requires the proteomes of the species under study in order to infer orthogroups. In this example we downloaded the proteomes sequences from [NCBI ftp site](ftp://ftp.ncbi.nlm.nih.gov/) using the following version:

- **GCF_000001735.3_TAIR10** for *A. thaliana* - [Download (11.3 Mb)](ftp://ftp.ncbi.nlm.nih.gov/genomes/refseq/plant/Arabidopsis_thaliana/latest_assembly_versions/GCF_000001735.3_TAIR10/GCF_000001735.3_TAIR10_protein.faa.gz),
- **GCF_000004255.1_v.1.0** for *A. lyrata* - [Download (7.0 Mb)](ftp://ftp.ncbi.nlm.nih.gov/genomes/refseq/plant/Arabidopsis_lyrata/all_assembly_versions/GCF_000004255.1_v.1.0/GCF_000004255.1_v.1.0_protein.faa.gz)

The samples are in the `exdata` folder.

We use this **bash** lines in order to produce orthogroups:

```{bash, eval = FALSE}
## blast everything
python OrthoFinder-master/orthofinder.py -f path/to/proteomes -t 3

## make orthogroups
python OrthoFinder-master/orthofinder.py -b path/to/results/WorkingDirectory
```

## Load orthogroups

The orthogroups are provided in the `OrthologousGroups.csv` file in Orthofinder's output.
Orthofinder provides both an `OrthologousGroups.csv` file and `OrthologousGroups.txt` file. 

The function `parse_orthogroups()` takes as input the path to the `OrthologousGroups.csv` file that you want to parse and load in R. Please note that `parse_orthogroups()` is designed for parsing **exclusively the *.csv* file** and not the *.txt* one.

```{r}
ogroups_path <- system.file("exdata",
                            "ogroups_sample.csv",
                            package = "crossr")
ogroups <- parse_orthogroups(ogroups_path)
```

The number of genes for orthogroup can be explored with the function `explore_ogroups()`.

```{r, fig.width= 7, fig.height=4, fig.cap="Genes per orthogroup in the sample dataset."}
explore_ogroups(ogroups)
```


# Convert protein ID in transcript ID

Since orthogroups are inferred on protein sequences but short reads are mapped on transcript sequences you might have to convert the **protein ID** in the orthogroups into **transcript ID**. 

The function `switch_ids()` performs this task. It requires an external file that maps the protein ID to transcript ID, which must be loaded in R.

In this case the **feature tables** containing the mapping were downloaded from **NCBI** from the same folder as the proteomes. Again, the samples are in the `exdata` folder.

Loading the feature table from refseq can be challenging, and the issues encountered are often table specific. I my own experiences one of the main sources of issues can be special characters in gene names and gene annotation within the table, such as hashes ( \# ) and quotes ( ' ). The quotes can be find pretty often because they are used in the 5' and 3' notation.

This code loads the *A. thaliana* feature table.
```{r}
thaliana_id_path <- system.file("exdata",
                                "thaliana_features_sample.txt",
                                package = "crossr")
    

thaliana_ids <- read.table(file = thaliana_id_path,
                           sep = "\t", header = FALSE,
                           stringsAsFactors = FALSE,
                           quote = "")

## Not sure why the header is commented with "#"
colnames(thaliana_ids) <-  scan(file = thaliana_id_path,
                                what = "",
                                nlines = 1)[-1]  
```

And this loads the *A. lyrata* one.
```{r}
lyrata_id_path <- system.file("exdata",
                              "lyrata_features_sample.txt",
                              package = "crossr")
    
lyrata_ids <- read.table(file = lyrata_id_path,
                         sep = "\t", header = FALSE,
                         stringsAsFactors = FALSE,
                         quote = "",
                         skip = 1,
                         comment.char = "") # One gene name contains "#"

## Not sure why the header is commented with "#"
colnames(lyrata_ids) <-  scan(file = lyrata_id_path,
                                what = "",
                                nlines = 1)[-1]  
```

Then we can use `switch_ids()` to convert the IDS from *A. thaliana* first, 

```{r}
ogroups <- switch_ids(ogroups = ogroups,
                      ids_table = thaliana_ids,
                      px_id = "product_accession",
                      tx_id = "related_accession",
                      mc.cores = 1)
```

and then the ones from *A. lyrata*.

```{r}
ogroups <- switch_ids(ogroups = ogroups,
                      ids_table = lyrata_ids,
                      px_id = "product_accession",
                      tx_id = "related_accession",
                      mc.cores = 1)
```

Note that the orthogroups contain IDS from **both species**, therefore `switch_ids()` must be called twice.

# Implement a S4 object

Now that we have loaded the expression data and the orthogroup data, we can start implementing an S4 element to organize and link together the expression sets, annotations and statistics that we have loaded and/or that we are going to produce.

We suggest [this chapter](http://adv-r.had.co.nz/S4.html) from Hadley Wickham's ***Advanced R*** book for an introduction to S4 objects in R.

## The ogset class

The *crossr* package defines the S4 class `ogset` which can be created with the `make_ogset` constructor function.

```{r}
og_set <-  make_ogset(og = ogroups,
                       spec1_exp = thaliana,
                       spec2_exp = lyrata)
```

The `ogset` class is designed to store orthogroup expression data and metadata; it has slots for orthogroup expression data, list of genes in orthogroups, single species expression data, design and annotation data. None of these elements is necessary when initializing the object with `make_ogset()` and also an empty set can be initialized.

The `ogset` class is modeled on Bioconductor Summarized experiment class.

When the `ogset` element is initialized, `make_ogset()` calls the function `check_ogset()` in order to check that the elements supplied match with each others. In details, `check_ogset()` checks that:

- the names of the orthogroups (`og`) match the rownames of the orthogroup expression set `og_eset` (if any is given),
- the row names of the annotations `og_annos` (if any) are contained in the names of the orthogroup list `og`,
- the row names of `colData` (if any) match the column names of the orthogroup expression set `og_eset`,
- the row names of `rowData` (if any) match the row names of the orthogroup expression set `og_eset`,
- the variable in `design` (if any) are colnames of `colData`,
- the rownames of `spec1_colData` match the colnames `spec1_exp` (if any),
- the rownames of `spec2_colData` match the colnames `spec2_exp` (if any),
- `exp_cond` is a character string,
- `exp_cond` is contained in columns of `spec1_colData` and `spec2_colData` and that those columns match.

`check_ogset()` can be called again at any time later. For example, after adding new elements to an `ogset` object, one can use it to check that the object is still valid. In case `check_ogset()` will return `TRUE`.

```{r}
check_ogset(og_set)
```


 
## Store sample information (colData) in the ogset class element

The **sample metadata** for the two species can be saved in the `spec1_colData` and `spec2_colData` slots of the `ogset` element.

In this case we have already expressed simplified information on the samples in the colnames of the two datasets. We can use this to specify very basic colData. Anyway, any kind of detailed sample metadata can be provided as `data.frame`.

```{r}
colnames(og_set@spec1_exp)
colnames(og_set@spec2_exp)

coldata_spec1 <- sapply(colnames(og_set@spec1_exp), function(i) strsplit(i, split = "_")[[1]][2])
coldata_spec1 <- data.frame(treat = coldata_spec1, stringsAsFactors = FALSE)
coldata_spec1

coldata_spec2 <- sapply(colnames(og_set@spec2_exp), function(i) strsplit(i, split = "_")[[1]][2])
coldata_spec2 <- data.frame(treat = coldata_spec2, stringsAsFactors = FALSE)
coldata_spec2

og_set@spec1_colData <- coldata_spec1
og_set@spec2_colData <- coldata_spec2
``` 

## Experimental factor

We have to provide information on which column of the colData datasets contains the main information on the **experimental conditions** under which we are **comparing** the two species (it could be different growth stages, treatments etc.).

The **name of the column** that encodes for the experimental factor in colData can be stored in the `exp_cond` slot.

This column must contain the same categorical variables in both the `spec1_colData` and the `spec2_colData` datasets; indeed, we are testing for the response to the same experimental condition in the two species.

```{r}
og_set@exp_cond <- "treat"
```

Afterwards, we can check that the ogset element is still valid.

```{r}
check_ogset(og_set)
```


# Collapse orthologues

As seen in *Figure 1* most of the orthogroup contain only two genes, generally one from *A. lyrata* and one from *A. thaliana*.

Many other orthogroups though contain different (any) number of genes from one species and from the other. A "quick and dirty"" solution to restore a one to one relationship among features is to collapse the orthologues by adding up the expression values of genes within the same orthogroup.

The basic **assumption** of this method is that **genes in the same orthogroup have the same function**.

The function `collapse_orthologs()` performs this task on an `ogset` class element, and returns the same element with the collapsed orthogroup expression set in the designated slot.

```{r}
og_set <- collapse_orthologs(og_set, mc.cores = 1)
```

Not all orthogroups contain genes from both species, the orthogroups that contain genes and hence expression data from both species are stored in the slot `og_exp` for further analysis,

```{r}
str(og_set@og_exp)
```

the ones that does not are stored in the slot `og_nomatch`

```{r}
str(og_set@og_nomatch)
```

After we have collapsed orthologues we can **check for consistency** of the `ogset` object with the function `check_ogset()`.

```{r}
check_ogset(og_set)
```

# Functional annotation for orthogroups

In order to get rudimentary functional annotations for the orthogroups we simply associate to an orthogroup all the functional annotations of the genes composing it.

In this example we use exclusively functional annotations of *A. thaliana*, since they are supposedly better than the annotations of *a. lyrata*.

## Functional annotations in practice

The feature names in the **feature table** of *A. thaliana* contain enough functional information for our purpose. This dataset was loaded in R under section 3 of this document in the object `thaliana_ids`.

The transcript and protein ids are in the `product_accession` column of the `thaliana_ids` dataframe,

```{r}
head(thaliana_ids$product_accession)
```

while the functional annotations are in the `name` column.

```{r}
head(thaliana_ids$name)
```

We use this code to extract the annotations for orthogroups and save them in the `rowData` slot.

```{r}
get_annos <- function(ogroup) 
{
    annos <- thaliana_ids[thaliana_ids$product_accession %in% ogroup, "name"]
    annos <- paste(unique(annos), collapse = " -- ")
    return(annos)
}

og_set@og_annos <- data.frame(func_annos = vapply(og_set@og, get_annos, character(1)),
                              stringsAsFactors = FALSE)

head(og_set@og_annos)
```

Afterwards, as always, we can check for consistency of the og_set element.

```{r}
check_ogset(og_set)
```


# Select groups that behave differently between species

After moving to orthogroup-wise expression, we can search for orthogroups that behave differently in the two species.

## Experimental design with interaction factor

Since genes from the two species within the same orthogroup have different sequences, we **do not perform direct comparison** of orthogroup expression level between the two species. Instead, we use a linear model in order to detect **relative differences** of expression of each orthogroup between the two species across the different experimental condition.

We achieve this by setting a design formula with one term for the species (`spc` in this case) one term for the treatment / experimental factor (`treat` in this case) and then an interaction factor (`spc:treat` in this case). So in this case the design formula will be `~ spc + treat + spc:treat`. Than we use the F-value of the interaction term as ranking feature.

Also, since the data have gone through many transformation it is hard to make assumption on their distribution, therefore we do not extract a p-value from the linear model. We use the F-value on the interaction term as a ranking feature, in order to select the genes that seem to behave differently among the species.

## Orthogroup ranking in practice

The function that fits the linear model and performs ANOVA is called `add_fit()`

In order to call it:

1. first we have load information on the column of the orthogroup expression set `og_exp` into the `ogset` element,
2. then we have to specify an experimental design.

### Specify colData for orthogroup-wise expression

The column data were previously encoded in the colnames of the two expression sets from the two different species. Those names are preserved in the colnames of the ortholog expression set.

```{r}
colnames(og_set@og_exp)
```

So I organize those info in a dataframe:

```{r}
coldata <- data.frame(spc = sapply(colnames(og_set@og_exp),
                                   function(i) strsplit(i, split = "_")[[1]][1]),
                      treat = sapply(colnames(og_set@og_exp),
                                     function(i) strsplit(i, split = "_")[[1]][2]),
                      stringsAsFactors = FALSE)

coldata
```

And provide those information to the `ogset` element in the `colData` slot.

```{r}
og_set@colData <- coldata
check_ogset(og_set)
```

### Specify design

The experimental design can be specified in the `design` slot:

```{r}
og_set@design <- ~ 0 + spc + treat + spc:treat
check_ogset(og_set)
```

### Estimate F-values and rank

Now we can calculate the F-values with `add_fit()`. 

We would suggest calculate the F-value on **log scaled value**, this is more appropriate for studying relative changes. This could be done with the argument `log_scale = TRUE` to the `add_fit()`. In this way, the `add_fit()` function:
 
1. add 1 to the every observation in the orthogroup expression dataset,
2. log transforms the data,
3 estimates the F-values,
4. writes the F-values in the `stats` slot.

So, in this way the F-value are estimated on `log(data + 1)`.

```{r}
og_set <- add_fit(ogset = og_set, log_scale = TRUE)
str(og_set@stats)
```

The info on the `log_scale` parameter are stored in the `metadata` slot.

```{r}
og_set@metadata
```


# Explore results

`add_fit()` saves an the F-statistic for all the terms in the design formula for every gene

```{r}
head(og_set@stats)
```

As mentioned above, the F-statistic associated to the interaction term `spc:treat` can be used as a ranking feature. The orthogroups with the highest F-value are the one that display a more divergent behavior in the two species.

The function `get_top_tags()` extracts the ID of the orthogroups with the highest F-value.

```{r}
tt <- get_top_tags(ogset = og_set,
                   rank_stat = "spc:treat",
                   n =  10)
tt
```

`get_top_tags()` returns a named numeric vector containing both the id of the top orthogroups and their associated statistics.

We can extract the functional information for the top tags with:

```{r}
og_set@og_annos[names(tt), ]
```


## Plot orthogroup expression

The function `plot_all_stages` plots a stripchart of the expression of the supplied orthogroup.

It requires as input:

- the **name** of the orthogroup,
- the **ogset element** that contains the details for that orthogroup,
- the **species encoding** in the design formula (in this case `spc`)
- the **condition encoding** in the design formula (in this case `treat`)

The code below selects the highest differentially expressed group and plots it.

```{r, fig.height=4, fig.width=5}
top_group <- names(get_top_tags(og_set, "spc:treat", 1))
plot_all_stages(orthogroup = top_group,
                ogset = og_set,
                species = "spc",
                condition = "treat",
                ylab = "Log Scaled TPM")
```

This can be combined with faceting in order to plot the expression of multiple orthogroups. These lines of code select and plot the top 12 differentially expressed orthogroups in a 4X3 faceting

```{r, fig.height=9, fig.width=7, results='hide'}
tt <- names(get_top_tags(og_set, "spc:treat", 12))
par(mfrow = c(4, 3))
sapply(tt,
       plot_all_stages,
       ogset = og_set,
       species = "spc",
       condition = "treat",
       ylab = "Log Scaled TPM")
```

Functional annotation can be displayed in the plot subtitle with the arguments `use_annos` and `annos_col`.
If you plan to display functional annotation, we suggest you to increase the upper margin of the plot with the function `par()`.

```{r, fig.height=4.5, fig.width=5}
par(mar = c(3, 4, 7, 2))

top_group <- names(get_top_tags(og_set, "spc:treat", 1))
plot_all_stages(orthogroup = top_group,
                ogset = og_set,
                species = "spc",
                condition = "treat",
                use_annos = TRUE,
                annos_col = "func_annos",
                ylab = "Log Scaled TPM")

```
 
## Plot all the genes composing a specific orthogroup

The workflow of `crossr` relies on the assumption that **all the genes composing one orthogroup have the same function**. 

If this **useful working assumption** was always verified, the expression values of those genes could be added without generating artifacts. Unfortunately this is not the case.

Therefore we suggest that, before drawing any conclusion, the user should explore by plotting **also the expression pattern of the single genes** contained in the orthogroups that are differentially expressed.

This can be achieved with the function `plot_og_genes`, that prints the number and the ID of the genes composing the user supplied orthogroup and plot their expression.

```{r, fig.width=9}
par(mfrow = c(1, 2))
plot_og_genes(ogroup = top_group,
              ogset = og_set,
              ylab = "TPM")
```

The orthogroup `r top_group` contains only 2 genes, but we can expect orthogroups of different size and different ratios of genes from one species and from the other.

# Example, orthogroup OG0002779

```{r}
example_group <- "OG0002779"
# also OG0000214 makes a good example
```


Orthogroup `r example_group` contains contains 4 genes, 2 from each species.

The F-value associated to this group is `r og_set@stats[example_group, "spc:treat"]`, and it ranks `r which(names(get_top_tags(og_set, "spc:treat", Inf)) == example_group)` out of 500 groups.

The associated annotation is: `r og_set@og_annos[example_group, "func_annos"]`(which is not incredibly helpful).

Looks like it is behaving differently in response to heat in the 2 species;

```{r, fig.height=4, fig.width=5}
plot_all_stages(orthogroup = example_group,
                ogset = og_set,
                species = "spc",
                condition = "treat",
                ylab = "Log Scaled TPM")
```

and it contains 4 genes of which one is behaving peculiarly differently than the others.

```{r, fig.height=6, fig.width=6}
par(mfrow = c(2, 2))
plot_og_genes(ogroup = example_group,
              ogset = og_set,
              ylab = "TPM")
```


# Clusters

We rank orthogroups by F-value in order to extract orthogroups that behave differently in the two species. But in this way, we extract orthogroups that behave differently in any kind of way. 

After extracting orthogroups that behave differently, we cluster their expression values in order to group the ones that behave differently in a similar way.

We can use the function `get_top_tags()` in order to apply an arbitrary cutoff and to extract the top n differentially expressed orthogroups, in this case 50.

```{r}
n <- 50
tt <- get_top_tags(ogset = og_set,
                   rank = "spc:treat",
                   n = n)
```

And then we cluster the the orthogroups with the `pheatmap()` function from the pheatmap package (but feel free to substitute it with your favourite clustering function if you prefer).

```{r, fig.height=8, fig.width=13.5}
pheatmap::pheatmap(t(scale(t(og_set@og_exp[names(tt), ]))),
                   cluster_cols = FALSE,
                   labels_row = paste(names(tt), og_set@og_annos[names(tt), "func_annos"], sep = ", "),
                   cellheight = 10, cellwidth = 25)

```


# Known issues and limitations

Applying the *crossr* workflow helped me a lot when I had to analyze cross-species RNA-seq datasets and when I had to extract biological observations from those data.

Nevertheless *crossr* has know issues and limitations.

## Limitations

While writing *crossr* I favoured simplicity over statistical precision, therefore *crossr* relies on **explorative data analysis** methods and does not perform rigorous testing. 

Making reliable statistical tests on cross-species transcriptomic data is a complicated and way beyond my expertise. Please, do not take results on a low number of replicates (2, 3 or 5) as definitive. Consider validating your observations with more measurements.

## Issues

I have identified three issues with *crossr* workflow. I am currently learning how to deal with these issues.

- Transcript isoforms are treated in the same way as transcript from other genes in the same orthogroup,
- Log transformation before ranking the genes might not be the most reliable option available,
- The cutoff between differentially expressed genes and non differentially expressed genes is arbitrary.


# Session info

Here is the output of `devtools::session_info()` on the system on which this document was compiled:

```{r, echo=FALSE}
devtools::session_info()
```

